Topological Equivariant Artist Model for Visualization Library Architecture

### 2023-06-08 paper/main.tex
notes for what needs to be filled in
backup cause vscode is trying to eat my files

% bare_jrnl_compsoc, texV1.4b, 2015/08/26, Michael Shell
\documentclass[10pt,journal,compsoc]{IEEEtran}

% *** MISC UTILITY PACKAGES ***

\newcommand{\note}[1]{\textcolor{magenta}{#1}} 
\usepackage[nocompress]{cite}
\usepackage[]{footmisc}
%\usepackage[notref, notcite]{showkeys}
\usepackage{hyperref} % autoref
% *** GRAPHICS RELATED PACKAGES ***
\ifCLASSINFOpdf
   \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
   \graphicspath{{./figures/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
   \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
   \graphicspath{{./figures/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
   \DeclareGraphicsExtensions{.eps}
\fi

% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.

% *** MATH PACKAGES ***
%% with other math-related packages, you may want to disable it.
\usepackage{amsmath, amsthm, amsfonts,amssymb,eulervm,xspace, mathtools}
\usepackage{stmaryrd}%mapsfrom 
%\renewcommand{\restriction}{\mathord{\upharpoonright}} %restriction w/p space
\usepackage{mathrsfs} % math script fonts
\usepackage{relsize} %bigger
\usepackage{bm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{remark}
\newtheorem{example}{Example}[section]
% *** SPECIALIZED LIST PACKAGES ***

\usepackage{xcolor}
\usepackage{algorithmic}
\usepackage[utf8]{inputenc}
\usepackage{multicol}
\usepackage{tikz}
\usetikzlibrary{cd} % commutative diagrams
\usepackage{pgfplots} %numberline
\DeclareUnicodeCharacter{2212}{−}
\usepgfplotslibrary{groupplots,dateplot}
\usetikzlibrary{patterns,shapes.arrows}

\pgfplotsset{compat=newest}
\newtheorem{axiom}{Axiom}
\newtheorem{prop}{Proposition} %math?
\usepackage[switch]{lineno}
\renewcommand{\linenumberfont}{\normalfont\bfseries\small\color{lightgray}}
\usepackage[outputdir=../]{minted}
\setminted[python]{fontsize=\scriptsize, 
                   linenos,
                   numbersep=8pt, 
                   frame=lines,
                   autogobble,
                   framesep=3mm, 
                   breaklines=True} 
% *** ALIGNMENT PACKAGES ***
\usepackage{array}
\usepackage{multirow}
\usepackage{tabulary} 
% IEEEtran contains the IEEEeqnarray family of commands

% *** SUBFIGURE PACKAGES ***
\ifCLASSOPTIONcompsoc
  \usepackage[caption=false,font=footnotesize,labelfont=sf,textfont=sf]{subfig}
\else
  \usepackage[caption=false,font=footnotesize]{subfig}
\fi

@@

% *** PDF, URL AND HYPERLINK PACKAGES ***
\usepackage{xurl} 
%\usepackage{hyperref} breaks compilation?
% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

@@

\usepackage{notation} %notation conventions
% correct bad hyphenation here

@@

% make the title area
\maketitle

\IEEEpeerreviewmaketitle

\IEEEraisesectionheading{\section{Introduction}\label{sec:intro}}

\IEEEPARstart{V}isualization is a structure preserving mapping from data to graphic. One way of verifying that the structure of the data is present in the graphic is to verify that the data set can be distilled from the graph, i.e. that the values read off a bar chart are equivalent to the values used to construct the bar chart. Since structure preservation requires that the $data\leftrightarrow graphic$ be bidirectional, changes to either data or graphic are expected to propogate to the other side of the transformation\cite{kindlmannAlgebraicProcessVisualization2014}. In this paper, we propose a methodology for describing structure, verifying structure preservation, and specifying the conditions for constructing a structure preserving map between data and graphics. We use category theory, which is the math of structures and their relations, to explicitly specify what it means for a visualization mapping to preserve structure in an explicit, comprehensive, and consistent way. We encode the structure of the data and graphics in a uniform generalizable manner by using a mathematical structure from algebraic topology called a fiber bundle; doing so allows us to discuss everything from lines to complex ND spaces with the same interface.

@@

 \begin{definition}
$f$ is \textbf{equivariant} when $f(act(g,x)) = act(g,f(x))$ for all $g$ in $G$ and for all $x$ in $X$\cite{nlab:equivariant, pittsNominalSetsNames2013}.
 \end{definition}
In an equivariant transform, mapping the data after acting on it $f(act)$ is equivalent to acting on the data after mapping it $act(f)$ since $G$ acts on both. A homomorphism does not require that the input and output are of the same type; instead

@@

The notion that visual encodings should be \textit{homomorphic} was proposed by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987} in his specification of \textit{A Presentation Tool} and the idea that visual transforms are \textit{equivariant} underlies Kindlemann and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014}'s Algebraic Visualization Design. Kindlemann and Scheidegger propose three specific types of structure preservation: that the visualization should not change if the data representation (i.e. the data container) changes, that data maps unambiguously to visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}, and that changing the data should correspond to changes in the visualization in a perceptually significant manner. A uniform abstract data representation layer is introduced in \autoref{sec:atct:sheaves}, the expectation of unambiguous mappings is expressed in \autoref{sec:atct:xi}, and the the correspondence between data and visual changes is included in the definition of equivariance introduced in \autoref{sec:artist:equivariant:artist}. We define structure preservation both in \note{field types and topology}

\subsection{Field Type Structure} 
\label{sec:related-work:equivariance}
 Bertin \cite{bertinSemiologyGraphicsDiagrams2011} codified the expectation that a data value and its visual encoding are expected to have compatible structure. The structure on values has traditionally been described  using the measurement scales-i.e. nominal, ordinal, interval, ratio - which Steven's classified by their mathematical group structure\cite{stevensTheoryScalesMeasurement1946a} and others have formally expanded to include more types \cite{leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}. Generally the structure on each field of the dataset is preserved separately, i.e. each column in a table is mapped to a distinct encoding. For example, nominal measurements are permutable, meaning that they are acted on by elements of a permutation group. For the visual transformation to be equivariant, that nominal measurement must be mapped to a visual encoding that is also permutable, such as distinct marker shapes or color hues. We evaluate whether the encoding is homomorphic when the visual encoding does not have a group structure. For example, partial orders are not invertible; therefore a visual encoding of partially ordered data is structure preserving when the ordering is mapped to equivalently ordered visual encodings. Steven's measurement scale's and the structure preservation they define is the basis of Wilkenson's scales\cite{wilkinsonGrammarGraphics2005} in Grammar of Graphics and derivative libraries, and in this work we generalize the concept of fields to support a wider array of field types. This also allows us to specify what consistency means when creating different views of data from a single dataset, as discussed in Qu and Hullman \cite{hullmanKeeping2018}, specifically in terms of each field type separate from the overall structure.

@@

The connectivity and relative positioning of elements in a dataset is broadly its topology\cite{wilkinsonGrammarGraphics2005} and the expectation that it be be reflected in the visual representation is present in even the earliest maps \cite{friendlyBriefHistoryData2008} and in Bertin's assertion that the dimensionality of the data match the class of representation \cite{bertinSemiologyGraphicsDiagrams2011}. In this work, we propose that using the formalism of topology provides us a uniform way to express dimensionality such that we can generalize the expectation that the data and visual dimensionalty match past points, lines, and areas.

@@

A topology is a partitioning of a space into its subspaces in a way where those subspaces can be composed back into the space and the choice of partitions is how we encode the connectivity of the data that we are interested in. For example, the GHCN\cite{lawrimoreGlobalHistoricalClimatology2011} dataset in \autoref{fig:related-work:continuity:ktypes} is a complex spatio-temporal record of weather station data. One way of encoding that each record is independent is to encode the underlying topology as a set of 0D $\cdots$ points, one for each record. Each of these discrete points is also represented visually as a discrete marker in the scatter plot. We can view the timeseries at each station as a 1D $---$ continuous line since weather station data is sampled from natural phenomena. We encode this topological structure as intervals, .i.e [start date, end date], and sets of sub-intervals, and we visually represent this strucure as continuous line plots. A third way of viewing the data is as a sparse sampling of a 2D $\blacksquare$ geospatial grid, where the connectivity is modeled as a surface and sub regions of that surface. In the visualization, that connectivity is expressed as plotting the stations in the map, visually denoting that these are points in the continuous geographical space that is New York State. By using topology as the underlying encoding scheme, we are able to represent the structure of points, lines, surfaces, and more using the same abstractions.

Since topological continuity is generalizable, we propose that encoding topology as part of a dataset type and a visualization type allows us to explicitly check which types of structure a visualization algorithm can preserve. Visual algorithms assume the topology of their input data, as described in taxonomies of visualization algorithms Chi\cite{chiTaxonomyVisualizationTechniques2000} and by Troy and M\"{o}ller \cite{toryRethinkingVisualizationHighlevel2004}, but generally do not verify that input structure. For example, a \texttt{line} algorithm often does not have a way to query whether a list of (x,y) coordinates is the distinct rows, the time series, or the list of stations in \autoref{fig:related-work:continuity:ktypes}. While plotting the time series as a continuous line would be correct, it would be incorrect for a visualization to indicate that the distinct rows or stations are connected in a 1D continuous manner. Encoding the expected input structure would enable a visualization algorithm to reject data that does not explicitely state that it matches those assumptions. This structure can be encoded in a generalizable way by building on Butler's work on using a mathematical structure called fiber bundles as an abstract data representation in visualization \cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989} where he discusses how bundles can be used to encode data in a consistent manner for the topological and field structures that are common in visualization. We sketch out fiber bundles in \autoref{sec:atct:fiber-bundles}, but his work provides a thorough introduction to bundles for visualization practitioners.

@@

\section{Formal Properties of Data \& Graphics}
\label{sec:atct}
In this section, we propose a mathematical abstraction of the data input and graphic prerendered output. This mathematical abstraction has a robust language for expressing topology and fields; expresses how to verify that data continuity is preserved on subset, distributed, and streaming data representations; and formalizes the expectation of a correspondence between data and visual elements.

@@

%For example, in \autoref{fig:local_trivalizations}, the bundles $\dtotal_{0}=\dtotal\restriction_{[\epsilon^{-}, {\frac{2\pi}{5}}^{+}]}$ and $\dtotal_{1}=\dtotal\restriction_{[{\frac{2\pi}{5}}^{-}, \epsilon^{-}]}$ are local trivializations of an arbitrary bundle $\dtotal$. These local trivializations overlap on the intersections of the open subsets of each trivalization $[\epsilon^{-}, \epsilon^{+}]$ and $[{\frac{2\pi}{5}}^{-}, {\frac{2\pi}{5}}^{+}]$. A set of trivializations can be glued into a larger bundle by defining transition maps that describe how to align the fibers in the overlapping space such that the sections of the bundle remain continuous. 
\note{add transition map construction formula}

@@

%In a \textit{trivial bundle}, the fibers are identical $\dfiber = \dfiber_{k}$ for all points $\dbasepoint \in dbase$ such that all the transtion maps are also identical. For example, in the cylinder case the fibers in the overlappping regions of the trivalizations point in the same direction $\uparrow$ to illustrate that they are equal $\dfiber_0\restriction_{\openset_1\cap\openset_2} = \dfiber_{1}\restriction_{\openset_1\cap\openset_2}$. Because the fibers point in the same direction, the  transition maps at both intersections map the fiber to itself $\uparrow\rightarrow\uparrow$.

@@

\subsubsection{Topological Structure: Base Space \dbase}
\label{sec:atct:fb:base}
\note{base space of bundle for key/value pair, topology in general 'cause data + viz continuity in same way}
Munzner~\cite{munznerChDataAbstraction} describes a key-value semantics for data representation and abstraction, where datasets can be queried with different kinds of keys depending on the dataset type - locations for values of spatial fields, times for values of timeseries, or indexes for values in flat or multidimensional tables.
We propose that these different types of keys are instances of different underlying topological organization of the data, and that we can use the structure of the opensets in the topological space to encode the extent to which the keys can vary continuously: flat or multidimensional tables have discrete points as indices, timeseries have points on a line as indices and spatial fields (regardless of whether they are scalar or tensor fields) have points in a spatial domain as indices. Therefore we choose to encode the topological structure of the data as the \textcolor{base}{base space} of a fiber bundle. Because the base space of a fiber bundle is a quotient topology\cite{munkresElementsAlgebraicTopology1984}, it divides the topological space into the largest number of open sets such that $\pi$ remains a continuous function. This means that the topology can be defined to have a resolution equal to the number of indicies in a dataset such that the key (continuity)-value (data) pairing is always preserved.

Following from Spivaks cataegorical abstraction of a database \cite{spivakSimplicialDatabases2009,spivakDatabasesAreCategories2010}, we also propose that the structure of the data types be formally specified as the objects of a category.

@@

The standard construction of a category from a topological space is that it has open set objects $\opensetc$ and inclusion morphisms $\opensetc_i \xrightarrow{\iota} \opensetc_j$ such that $\opensetc_i \subseteq \opensetc_j$\cite{riehlCategoryTheoryContext}. The composability property expresses that inclusion is transitive, while associativity expresses that the inclusion functions can be curried in various equivalent groupings. By formally specifying the properties of the topological structure datatypes as $\mathcal{\dbase}$, we can express that these are the properties that are required as part of the implementation of the data type objects.

For example, the disjoint union of two bundles, as shown in \autoref{fig:atct:local_trivalizations}, is the coproduct $\dbase^{a} \subseteq_{\dbase^{c}} \dbase^{b}$ over an overlap $\dbase^{c}$ and therefore the inclusion morphism must be commmutative:
\begin{minipage}[c]{0.25\columnwidth} 
  \begin{equation}
    \label{eq:atct:base_union}
  \begin{tikzcd}
    \dbasec^{a}\sqcup_{\dbasec^{c}}\dbasec^{b} 
    & \dbasec_{b} \arrow[l, "\iota_b"', color=base] \\
    \dbasec_a \arrow[u, "\iota_a", color=base] 
    & \dbasec_c 
    \arrow[u, "\iota_c"', color=base] 
    \arrow[l, "\iota_c"', color=base] 
  \end{tikzcd}
\end{equation}
\end{minipage}\hfil
\begin{minipage}[c]{0.65\columnwidth}
  \begin{tikzpicture}[commutative diagrams/every diagram]
    \matrix[matrix of math nodes, name=m, commutative diagrams/every cell]{
      (0,1) & (.55,1) \\
      (0,.45) & (.40, 60)\\  
    };

@@

To build on the rich typing system provided by Spivak, we define the \textcolor{fiber}{fiber space} \dfiberc\ to be the space of all possible data records
\begin{equation}
  \dfiberc \coloneqq \{\delement: \fnames \rightarrow \mathscr{U}_{\sigma} \bigm{\vert} \pi_{\sigma}(\delement(\fname)) = \fname\;for\;all\; \fname \in \fnames \}
\end{equation}
such that the preimage of a point is the corresponding data type domain $\pi^{-1}(\dbasepoint) = \dfiber_{k} = \mathscr{U}_{{\sigma}_{\dbasepoint}}$. 
Adopting Spivak's fiber bundle construction of types allows our model to reuse types so long as the field names are distinct and that field values can be accesed by field name,  since those are sections on $\mathcal{U}_{\sigma}$. Furthermore, since domains $\mathscr{U}_{{\sigma}}$ of types are a mathematical space, multi-dimensional fields can be encoded in the same manner as single dimensional fields and fields can have different names but the same type.

As with the base space category $\mathcal{\dbase}$, we propose a fiber category $\mathcal{\dfiberc}$ to encapsulate the field types of the data. The fiber category has a single object $\dfiberc$ of an arbitrary type and morphisms on the fiber object $\dfunctc \in Hom(\dfiberc, \dfiberc)$. We can also equip the category with any operators or relations that are part of he mathematical structure of the field type. For example we can equip the category with a comparison operator, which is part of the definition of the monoidal structure of a partially ordered ranking variable \cite{bruggemannRankingPrioritizationMultiindicator2011} or the group structure of Steven's ordinal measurment scale \cite{stevensTheoryScalesMeasurement1946, leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}. Steven's other scales are summarized in \autoref{tab:appendix:summary:stevens}.

The fiber category $\mathcal{\dfiberc}$ is also equipped with a bifunctor because it is a monoidal category and this functor provides a method for combining fiber types. The bifunctor $\otimes: \mathcal{\dfiber} \times \mathcal{\dfiber} \rightarrow \mathcal{\dfiber}$ allows us to express fields that contain multityped values. For example, wind can be represented as two fields $\dfiber_{speed} \times \dfiber_{direction}$ or a composite fiber field $\dfiber_{speed} \otimes \dfiber_{direction} = \dfiber_{wind}$. The $\otimes$ encapsulates both the sets associated with each fiber $\mathbb{R} \times \mathbb{R} = \mathbb{R}^{2}$ and the morphisms associated with each functor $(\dfunct_{speed}, \dfunct_{direction}) = \dfunct_{wind}$. Combining fibers can be verified by checking that when a fiber component $\dfiber^{c}$ is present in both $\dfiber^{a}$ and $\dfiber^{b}$, it is identical when projected out of either such that the *product* diagram commutes:
\begin{minipage}

>>>>

\documentclass[journal]{IEEEtran}

% *** MISC UTILITY PACKAGES ***
\newcommand{\note}[1]{\textcolor{magenta}{#1}} 
\usepackage{silence}

\WarningsOff[fixltx2e]
\usepackage[switch]{lineno}
\renewcommand{\linenumberfont}{\normalfont\bfseries\small\color{lightgray}}

% references
\usepackage[nocompress]{cite}
\usepackage[]{footmisc}
\usepackage{bookmark}

% *** GRAPHICS RELATED PACKAGES ***
\usepackage{graphicx}
\graphicspath{{./figures/}}

% *** MATH PACKAGES ***
%% with other math-related packages, you may want to disable it.
\usepackage{amsmath, amsthm, amsfonts,amssymb,eulervm,xspace, mathtools}
\usepackage{stmaryrd}%mapsfrom 
%\renewcommand{\restriction}{\mathord{\upharpoonright}} %restriction w/p space
\usepackage{mathrsfs} % math script fonts
\usepackage{relsize} %bigger
\usepackage{bm}

% theorem environments
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{remark}
\newtheorem{example}{Example}[section]
\newtheorem{axiom}{Axiom}
\newtheorem{prop}{Proposition}

% *** diagrams
\usepackage{tikz}
\usetikzlibrary{cd} % commutative diagrams
\usepackage{pgfplots} %numberline
\DeclareUnicodeCharacter{2212}{−}
\usepgfplotslibrary{groupplots,dateplot}
\usetikzlibrary{patterns,shapes.arrows}
\pgfplotsset{compat=newest}

% *** SPECIALIZED LIST PACKAGES ***
\usepackage{xcolor}
\usepackage[utf8]{inputenc}

\usepackage[outputdir=.]{minted}
\setminted[python]{fontsize=\scriptsize, 
                   linenos,
                   numbersep=8pt, 
                   frame=lines,
                   autogobble,
                   framesep=3mm, 
                   breaklines=True} 
% *** ALIGNMENT PACKAGES ***
\usepackage{multicol}
\usepackage{array}
\usepackage{multirow}
\usepackage{tabulary} 
% IEEEtran contains the IEEEeqnarray family of commands

% *** SUBFIGURE PACKAGES ***
\usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}

@@

% *** PDF, URL AND HYPERLINK PACKAGES ***
\usepackage{xurl}

@@

\usepackage{notation} %notation conventions
% correct bad hyphenation here
%\hyphenation{}

@@

% make the title area
\maketitle
\IEEEpeerreviewmaketitle
\IEEEraisesectionheading{\section{Introduction}\label{sec:intro}}

\IEEEPARstart{V}{isualization} is a structure preserving mapping from data to graphic. One way of verifying that the structure of the data is present in the graphic is to verify that the data set can be distilled from the graph, i.e. that the values read off a bar chart are equivalent to the values used to construct the bar chart. Since structure preservation requires that the $data\leftrightarrow graphic$ be bidirectional, changes to either data or graphic are expected to propagate to the other side of the transformation\cite{kindlmannAlgebraicProcessVisualization2014}. In this paper, we propose a methodology for describing structure, verifying structure preservation, and specifying the conditions for constructing a structure preserving map between data and graphics. We use category theory, which is the math of structures and their relations, to explicitly specify what it means for a visualization mapping to preserve structure in an explicit, comprehensive, and consistent way. We encode the structure of the data and graphics in a uniform generalizable manner by using a mathematical structure from algebraic topology called a fiber bundle; doing so allows us to discuss everything from lines to complex ND spaces with the same interface.

@@

 \begin{definition}
$f$ is \textbf{equivariant} when $f(act(g,x)) = act(g,f(x))$ for all $g$ in $G$ and for all $x$ in $X$\cite{nlab:equivariant,pittsNominalSetsNames2013}.
 \end{definition}
In an equivariant transform, mapping the data after acting on it $f(act)$ is equivalent to acting on the data after mapping it $act(f)$ since $G$ acts on both. A homomorphism does not require that the input and output are of the same type; instead

@@

The notion that visual encoding functions should be \textit{homomorphisms} was proposed by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987} in his specification of \textit{A Presentation Tool} and the idea that visual transforms are \textit{equivariant} underlies Kindlemann and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014}'s Algebraic Visualization Design. Kindlemann and Scheidegger propose three specific types of structure preservation: that the visualization should not change if the data representation (i.e. the data container) changes, that data maps unambiguously to visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}, and that changing the data should correspond to changes in the visualization in a perceptually significant manner. A uniform abstract data representation layer is introduced in \autoref{sec:atct:sheaves}, the expectation of unambiguous mappings is expressed in \autoref{sec:atct:xi}, and the the correspondence between data and visual changes is included in the definition of equivariance introduced in \autoref{sec:artist:equivariant:artist}. We define structure preservation both in \note{field types and topology}

\subsection{Field Type Structure} 
\label{sec:related-work:equivariance}
 Bertin \cite{bertinSemiologyGraphicsDiagrams2011} codified the expectation that a data value and its visual encoding are expected to have compatible structure. The structure on values has traditionally been described  using the measurement scales-i.e. nominal, ordinal, interval, ratio - which Steven's classified by their mathematical group structure\cite{stevensTheoryScalesMeasurement1946} and others have formally expanded to include more types \cite{leleaFormalizationMeasurementScale1971 thomasMathematizationNotMeasurement2014}. Generally the structure on each field of the dataset is preserved separately, i.e. each column in a table is mapped to a distinct encoding. For example, nominal measurements are permutable, meaning that they are acted on by elements of a permutation group. For the visual transformation to be equivariant, that nominal measurement must be mapped to a visual encoding that is also permutable, such as distinct marker shapes or color hues. We evaluate whether the encoding is a homomorphism when the visual encoding does not have a group structure. For example, partial orders are not invertible; therefore a visual encoding of partially ordered data is structure preserving when the ordering is mapped to equivalently ordered visual encodings. Steven's measurement scale's and the structure preservation they define is the basis of Wilkenson's scales\cite{wilkinsonGrammarGraphics2005} in Grammar of Graphics and derivative libraries, and in this work we generalize the concept of fields to support a wider array of field types. This also allows us to specify what consistency means when creating different views of data from a single dataset, as discussed in Qu and Hullman \cite{hullmanKeeping2018}, specifically in terms of each field type separate from the overall structure.

@@

The connectivity and relative positioning of elements in a dataset is broadly its topology\cite{wilkinsonGrammarGraphics2005} and the expectation that it be be reflected in the visual representation is present in even the earliest maps \cite{friendlyBriefHistoryData2008} and in Bertin's assertion that the dimensionality of the data match the class of representation \cite{bertinSemiologyGraphicsDiagrams2011}. In this work, we propose that using the formalism of topology provides us a uniform way to express dimensionality such that we can generalize the expectation that the data and visual dimensionality match past points, lines, and areas.

@@

A topology is a partitioning of a space into its subspaces in a way where those subspaces can be composed back into the space and the choice of partitions is how we encode the connectivity of the data that we are interested in. For example, the GHCN\cite{lawrimoreGlobalHistoricalClimatology2011} dataset in \autoref{fig:related-work:continuity:ktypes} is a complex spatiotemporal record of weather station data. One way of encoding that each record is independent is to encode the underlying topology as a set of 0D $\cdots$ points, one for each record. Each of these discrete points is also represented visually as a discrete marker in the scatter plot. We can view the time series at each station as a 1D $---$ continuous line since weather station data is sampled from natural phenomena. We encode this topological structure as intervals, .i.e [start date, end date], and sets of sub-intervals, and we visually represent this structure as continuous line plots. A third way of viewing the data is as a sparse sampling of a 2D $\blacksquare$ geospatial grid, where the connectivity is modeled as a surface and sub regions of that surface. In the visualization, that connectivity is expressed as plotting the stations in the map, visually denoting that these are points in the continuous geographical space that is New York State. By using topology as the underlying encoding scheme, we are able to represent the structure of points, lines, surfaces, and more using the same abstractions.

Since topological continuity is generalizable, we propose that encoding topology as part of a dataset type and a visualization type allows us to explicitly check which types of structure a visualization algorithm can preserve. Visual algorithms assume the topology of their input data, as described in taxonomies of visualization algorithms Chi\cite{chiTaxonomyVisualizationTechniques2000} and by Troy and M\"{o}ller \cite{toryRethinkingVisualizationHighlevel2004}, but generally do not verify that input structure. For example, a \texttt{line} algorithm often does not have a way to query whether a list of (x,y) coordinates is the distinct rows, the time series, or the list of stations in \autoref{fig:related-work:continuity:ktypes}. While plotting the time series as a continuous line would be correct, it would be incorrect for a visualization to indicate that the distinct rows or stations are connected in a 1D continuous manner. Encoding the expected input structure would enable a visualization algorithm to reject data that does not explicitly state that it matches those assumptions. This structure can be encoded in a generalizable way by building on Butler's work on using a mathematical structure called fiber bundles as an abstract data representation in visualization \cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989} where he discusses how bundles can be used to encode data in a consistent manner for the topological and field structures that are common in visualization. We sketch out fiber bundles in \autoref{sec:atct:fiber-bundles}, but his work provides a thorough introduction to bundles for visualization practitioners.

@@

\section{Formal Properties of Data \& Graphics}
\label{sec:atct}
In this section, we propose a mathematical abstraction of the data input and graphic pre-rendered output. This mathematical abstraction has a robust language for expressing topology and fields; expresses how to verify that data continuity is preserved on subset, distributed, and streaming data representations; and formalizes the expectation of a correspondence between data and visual elements.

@@

%For example, in \autoref{fig:local_trivalizations}, the bundles $\dtotal_{0}=\dtotal\restriction_{[\epsilon^{-}, {\frac{2\pi}{5}}^{+}]}$ and $\dtotal_{1}=\dtotal\restriction_{[{\frac{2\pi}{5}}^{-}, \epsilon^{-}]}$ are local trivializations of an arbitrary bundle $\dtotal$. These local trivializations overlap on the intersections of the open subsets of each trivialization $[\epsilon^{-}, \epsilon^{+}]$ and $[{\frac{2\pi}{5}}^{-}, {\frac{2\pi}{5}}^{+}]$. A set of trivializations can be glued into a larger bundle by defining transition maps that describe how to align the fibers in the overlapping space such that the sections of the bundle remain continuous. 
\note{add transition map construction formula}

@@

%In a \textit{trivial bundle}, the fibers are identical $\dfiber = \dfiber_{k}$ for all points $\dbasepoint \in dbase$ such that all the transition maps are also identical. For example, in the cylinder case the fibers in the overlapping regions of the trivializations point in the same direction $\uparrow$ to illustrate that they are equal $\dfiber_0\restriction_{\openset_1\cap\openset_2} = \dfiber_{1}\restriction_{\openset_1\cap\openset_2}$. Because the fibers point in the same direction, the  transition maps at both intersections map the fiber to itself $\uparrow\rightarrow\uparrow$.

@@

\subsubsection{Topological Structure: Base Space \dbase}
\label{sec:atct:fb:base}
\note{base space of bundle for key/value pair, topology in general 'cause data + viz continuity in same way}
Munzner~\cite{munznerChDataAbstraction} describes a key-value semantics for data representation and abstraction, where datasets can be queried with different kinds of keys depending on the dataset type - locations for values of spatial fields, times for values of time series, or indexes for values in flat or multidimensional tables.
We propose that these different types of keys are instances of different underlying topological organization of the data, and that we can use the structure of the opensets in the topological space to encode the extent to which the keys can vary continuously: flat or multidimensional tables have discrete points as indices, time series have points on a line as indices and spatial fields (regardless of whether they are scalar or tensor fields) have points in a spatial domain as indices. Therefore we choose to encode the topological structure of the data as the \textcolor{base}{base space} of a fiber bundle. Because the base space of a fiber bundle is a quotient topology\cite{munkresElementsAlgebraicTopology1984}, it divides the topological space into the largest number of open sets such that $\pi$ remains a continuous function. This means that the topology can be defined to have a resolution equal to the number of indices in a dataset such that the key (continuity)-value (data) pairing is always preserved.

Following from Spivak's categorical abstraction of a database \cite{spivakSimplicialDatabases2009,spivakDatabasesAreCategories2010}, we also propose that the structure of the data types be formally specified as the objects of a category.

@@

The standard construction of a category from a topological space is that it has open set objects $\opensetc$ and inclusion morphisms $\opensetc_i \xrightarrow{\iota} \opensetc_j$ such that $\opensetc_i \subseteq \opensetc_j$\cite{riehlCategoryTheoryContext}. The composability property expresses that inclusion is transitive, while associativity expresses that the inclusion functions can be curried in various equivalent groupings. By formally specifying the properties of the topological structure data types as $\mathcal{\dbase}$, we can express that these are the properties that are required as part of the implementation of the data type objects.

For example, the disjoint union of two bundles, as shown in \autoref{fig:atct:local_trivalizations}, is the coproduct $\dbase^{a} \subseteq_{\dbase^{c}} \dbase^{b}$ over an overlap $\dbase^{c}$ and therefore the inclusion morphism must be commutative:
\begin{minipage}[c]{0.25\columnwidth} 
  \begin{equation}
    \label{eq:atct:base_union}
  \begin{tikzcd}
    \dbasec^{a}\sqcup_{\dbasec^{c}}\dbasec^{b} 
    & \dbasec_{b} \arrow[l, "\iota_b"', color=base] \\
    \dbasec_a \arrow[u, "\iota_a", color=base] 
    & \dbasec_c 
    \arrow[u, "\iota_c"', color=base] 
    \arrow[l, "\iota_c"', color=base] 
  \end{tikzcd}
\end{equation}
\end{minipage}\hfill

\begin{minipage}[c]{0.65\columnwidth}
  \note{will be replaced by lines showing intervals}
  \begin{tikzpicture}[commutative diagrams/every diagram]
    \matrix[matrix of math nodes, name=m, commutative diagrams/every cell]{
      (0,1) & (.55,1) \\
      (0,.45) & (.40, 60)\\  
    };

@@

To build on the rich typing system provided by Spivak, we define the \textcolor{fiber}{fiber space} \dfiberc\ to be the space of all possible data records
\begin{equation}
  \dfiberc \coloneqq \{\delement: \fnames \rightarrow \mathscr{U}_{\sigma} \bigm{\vert} \pi_{\sigma}(\delement(\fname)) = \fname\;for\;all\; \fname \in \fnames \}
\end{equation}
such that the preimage of a point is the corresponding data type domain $\pi^{-1}(\dbasepoint) = \dfiber_{k} = \mathscr{U}_{{\sigma}_{\dbasepoint}}$. 
Adopting Spivak's fiber bundle construction of types allows our model to reuse types so long as the field names are distinct and that field values can be accessed by field name,  since those are sections on $\mathcal{U}_{\sigma}$. Furthermore, since domains $\mathscr{U}_{{\sigma}}$ of types are a mathematical space, multi-dimensional fields can be encoded in the same manner as single dimensional fields and fields can have different names but the same type.

As with the base space category $\mathcal{\dbase}$, we propose a fiber category $\mathcal{\dfiberc}$ to encapsulate the field types of the data. The fiber category has a single object $\dfiberc$ of an arbitrary type and morphisms on the fiber object $\dfunctc \in Hom(\dfiberc, \dfiberc)$. We can also equip the category with any operators or relations that are part of he mathematical structure of the field type. For example we can equip the category with a comparison operator, which is part of the definition of the monoidal structure of a partially ordered ranking variable \cite{bruggemannRankingPrioritizationMultiindicator2011} or the group structure of Steven's ordinal measurement scale \cite{stevensTheoryScalesMeasurement1946, leaformalizationmeasleaFormalizationMeasurementScale1971, thomasMathematizationNotMeasurement2014}. Steven's other scales are summarized in \autoref{tab:appendix:summary:stevens}.

The fiber category $\mathcal{\dfiberc}$ is also equipped with a bifunctor because it is a monoidal category and this functor provides a method for combining fiber types. The bifunctor $\otimes: \mathcal{\dfiber} \times \mathcal{\dfiber} \rightarrow \mathcal{\dfiber}$ allows us to express fields that contain complexly typed values. For example, wind can be represented as two fields $\dfiber_{speed} \times \dfiber_{direction}$ or a composite fiber field $\dfiber_{speed} \otimes \dfiber_{direction} = \dfiber_{wind}$. The $\otimes$ encapsulates both the sets associated with each fiber $\mathbb{R} \times \mathbb{R} = \mathbb{R}^{2}$ and the morphisms associated with each functor $(\dfunct_{speed}, \dfunct_{direction}) = \dfunct_{wind}$. Combining fibers can be verified by checking that when a fiber component $\dfiber^{c}$ is present in both $\dfiber^{a}$ and $\dfiber^{b}$, it is identical when projected out of either such that the *product* diagram commutes:
